A Large-scale Study of Polymorphism in R

### 2019-04-02 paper/main.tex
a bit on L2

L0 is limited in expressivity. To address its limitations, we augment it
scalars, parametrized lists.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.

@@

\subsection{L2 Types}
 {\bf L2}: The third level adds attributes lists...

While we do consider some class and attribute information when constructing
our types, it is also valuable to consider them entirely separate from
types.  Users are free to define their own classes, and arbitrarily add and
remove attributes, and interesting patterns might arise in considering them.
That said, not attribute patterns and classes are necessarily interesting,
and indeed some of R's built-in data structures use them to some extend.
For one, data frames have a {\tt data.frame} class (explicitly on the class
attribute).  Other standard patterns are:

\begin{itemize}
\item data frames have three attributes: {\tt names} for column names, {\tt
  row.names} for row names, and {\tt class} (which is instantiated to {\tt
  "data.frame"} upon creation).
\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
  list with a {\tt dim} attribute {\it is} a matrix).
\item lists and vectors may have a {\tt names} attribute, which assigns
  names to locations.
\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \texttt{"factor"}.
\end{itemize}

@@

At L0, the type system is aligned with the types of the core language.  We
have observed over 173K monomorphic functions; this accounts for 80\% of our
corpus.  In terms of argument positions, 92\% of arguments are monomorphic.

Table~\ref{tab:L0argcounts} shows the polymorphicity of argument positions
 in R function.  From there, we build up a picture of the polymorphicity of
 functions as a whole, and the data for this function polymorphism can be
 found in Figure~\ref{tab:L0funcounts}.

\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
Type                   & Count & Percentage \\  \hline
%  Full Monomorphic & 785K & 85.87 \\ 
  Monomorphic in Type & 866K & 94.76 \\ 
%  Monomorphic in Class & 827K & 90.41 \\ 
%  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

\begin{table}[ht]
\label{tab:L0funcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\
  \hline
%  Full Monomorphic & 145K & 66.77 \\ 
  Monomorphic in Type & 186K & 85.78 \\ 
%  Monomorphic in Class & 163K & 75.21 \\
%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Function polymorphism in L0.}
\end{table}

From Table~\ref{tab:L0argcounts}, we glean that argument positions aren't
very polymorphic in L0.  An argument which is "Monomorphic in Type" is
monomorphic w.r.t. \code{typeof}, in that it was only ever observed to have
one \code{typeof}.  Similarly, an argument is "Monomorphic in Class" if it
is only ever called with one {\tt class}.  As we said earlier, attributes
are a little more complicated, and an argument which is "Monomorphic in
Attribute Pattern" was only ever observed to have one attribute pattern (and
that pattern may itself contain many attributes).

As for the polymorphism of functions as a whole, we glean from
Figure~\ref{tab:L0funcounts} that a sizeable number of functions are
polymorphic.  Functions are said to be polymorphic if at least one of their
arguments or return are polymorphic.  \AT{Say more?}

At present, we will only concern ourselves with \code{typeof} polymorphism,
and we will later worry about classes and attributes.  To get an idea of
what kinds of polymorphism occur, we will look at the most commonly
occurring polymorphic argument signatures.  Table~\ref{tab:L0toppoly} has
the top polymorphic signatures for L0.

@@

Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double,
integer polymorphism is unsurprising), others are strange (e.g., character,
double), but in fact they all have some explanation.

\begin{enumerate}
\item \{\D, \I\}: in R, doubles and integers are effectively
  interchangeable, as casting from one to the other is straightforward and
  automatically performed when appropriate.  For example, \code{c(1, 2)[1.9]
    == 1}.  This distinction is mainly just relevant to the implementation,
  as integers and doubles are stored and dealt with differently in the
  runtime but programmers only see the distinction when printing.

\item \{\D, \l\} and \{\C, \l\}: at first glance, these type signatures are
  suspect.  Why would so many programmers want to pass a double or a list to
  some function?  The key insight here is that doubles and characters are
  {\it vectorized} in R, meaning that a double is really a vector of
  doubles.  Thus, these seemingly strange signatures are revealed to be
  entirely reasonable: these signatures are likely to describe arguments
  which are either vectors or lists.

\item \{\ANY,\sN\}: in most languages, NULL is an inhabitant of all types,
  and we see that this is not the case in R.  That said, there is a pattern
  of many functions having default or optional arguments, and in many of
  these cases NULL is the default value.

\item \{\D,\L\}: the first values of type logical that come to mind are TRUE
  and FALSE, which does little to explain this strange signature.  Indeed,
  the key here is an implementation detail of R's: recall that NAs inhabit
  every type.  As it happens, the default type of NA is logical (e.g.,
  \code{typeof(NA) == "logical"}), and this is likely what is being seen
  here.  We will control for this by distinguishing NAs with their own type,
  {\tt raw\_NA}, but we will get to that in time.

\item \{\C,\D\}, \{\C,\I\} and \{character, double, integer\}: yet more
  strange signatures.  These signatures are a little more subtle to explain.
  For one, indexing lists and vectors can be done with either strings (i.e.,
  to access a named index) or with numbers.  Another option is dates: R has
  many ways to present dates, and strings and numbers feature among them.
  Finally, many \AT{get numbers} functions in R return a string if an error
  occurs during execution.

\end{enumerate}

How will we deal with this?

\begin{enumerate}
\item consolidating doubles and integers into a single real-number type
  would account for all of this polymorphism.
\item to get an idea of how often vectors and lists co-occur in signatures,
  we need to expand our analysis of vectorized types.  Notably, when we
  encounter e.g. a double, we should collect length information to see if
  the double is a vector of doubles or just a scalar.  This will feature in
  our next type system, L1, which we will get to momentarily.
\item if we roll NULLs into every type, we can cleanly deal with these X,
  NULL signatures.  As we have established that an annotation for an
  optional argument is useful, this will allow us to reveal other
  interesting patterns.
\item similarly with NULLs, we're interested in rolling NAs in to all types.
  As we mentioned, this is actually what the R runtime does, but a default
  type needed to be chosen for NAs appearing without context.  Subsuming NAs
  should help to reveal other interesting patterns.
\item dealing with this polymorphism will be more tricky.  We might need to
  pull class information in (to pick up on dates), and possibly make a
  signature judgment depending on which argument position has the signature
  (e.g. a \{character, X\} signature on a return).  We will come back to
  this later.

\end{enumerate}

There are some clear limitations to the L0 type system that we should
address before working through any of these other issues As we mentioned,
\code{typeof} makes no distinction between vectors and scalars, as indeed
there is no such distinction in R.  But as we saw, we might benefit from
such a distinction, and indeed an all-new runtime environment design for R
might benefit from e.g. not needing to vectorize scalar values.  This is one
of the many modifications we make, all of which are thoroughly explored in
the following section.
%
%
%
%
\subsection{L1}

@@

\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \sD & numeric & \{\} \\
y & \D & numeric & \{\} \\
retv & list<double> & list & \{\xspace example: \sC\}
\end{tabular}
\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}

>>>>

L0 is limited in expressivity. To address its limitations, we augment it with
scalars and parametrized lists.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.

@@

\subsection{L2 Types}

Though more expressive than L0, L1 only incorporates a limited amount of
class and attribute information. For instance, L1 says nothing about user-defined
classes and non-standard attributes. The standard attribute patterns in R are:

\begin{itemize}
\item data frames have three attributes: {\tt names} for column names, {\tt
  row.names} for row names, and {\tt class} (which is instantiated to {\tt
  "data.frame"} upon creation).
\item matrices have at least a {\tt dim} attribute with matrix dimensions,
  and optionally a {\tt dimnames} attribute for dimension names. Curiously,
  the {\tt dim} attribute is all that is needed to define a matrix (i.e., a
  vector with a {\tt dim} attribute {\it is} a matrix).
\item lists and vectors may have a {\tt names} attribute, which assigns
  names to locations.
\item to ease dealing with large sorted sets of non-numeric values, R offers
  the \code{factor} and \code{levels} functions.  \code{factor(x,
    levels=someOrder)} adds a {\tt "levels"} attribute to \code{x}, which
  specifies an ordering for \code{x} according to the \code{sortedOrder}
  parameter.  \code{factor} also changes the class of the parameter being
  factored to \texttt{"factor"}.
\end{itemize}

All of the possible \code{typeof}s in correspond to a class, so we must control
for these as well.

Now, recall that the attributes of a value is a list of name, value pairs of metadata
on the value accessible through the \code{attributes} function. For our purposes,
we will only consider the names of attributes, and not values nor their types. This
avoids expensive recursion, and including type information in attribute patterns 
creates a lot of noise in the data that hides other interesting patterns.

@@

At L0, the type system is mostly aligned with the types of the core language.  We
have observed over 186K monomorphic functions; this accounts for 86\% of our
corpus.  In terms of argument positions, 95\% of arguments are monomorphic.

Table~\ref{tab:L0argcounts} shows the polymorphicity of argument positions
 in R function.  From there, we build up a picture of the polymorphicity of
 functions as a whole, and the data for this function polymorphism can be
 found in Table~\ref{tab:L0funcounts}.

\begin{table}[ht]\label{tab:L0argcounts}\centering\begin{tabular}{lrr}  \hline
Type                   & Count & Percentage \\  \hline
%  Full Monomorphic & 785K & 85.87 \\ 
  Monomorphic in Type & 870K & 95.2\% \\ 
%  Monomorphic in Class & 827K & 90.41 \\ 
%  Monomorphic in Attribute Pattern & 825K & 90.27 \\ 
  Total Seen & 914K & --- \\ 
   \hline
\end{tabular}
\caption{Argument position polymorphism in L0.}
\end{table}

\begin{table}[ht]
\label{tab:L0funcounts}
\centering
\begin{tabular}{lrr}
  \hline
Type & Count & Percentage \\
  \hline
%  Full Monomorphic & 145K & 66.77 \\ 
  Monomorphic in Type & 187K & 86.4\% \\ 
%  Monomorphic in Class & 163K & 75.21 \\
%  Monomorphic in Attribute Pattern & 161K & 74.37 \\
  Total Seen & 216K & --- \\
   \hline
\end{tabular}
\caption{Function polymorphism in L0.}
\end{table}

Tables~\ref{tab:L0argcounts} and~\ref{tab:L0funcounts} show that the overall amount of polymorphism in R is quite low, but that doesn't quite tell the whole story.
To get an idea of what kinds of polymorphism occur, we will look at the most commonly occurring polymorphic argument signatures.  
Table~\ref{tab:L0toppoly} has the top polymorphic signatures for L0.

@@

\AT{TODO explain results in table.}

%Some entries in the Table~\ref{tab:L0toppoly} are clear (e.g., double,
%integer polymorphism is unsurprising), others are strange (e.g., character,
%double), but in fact they all have some explanation.

%\begin{enumerate}
%\item \{\D, \I\}: in R, doubles and integers are effectively
%  interchangeable, as casting from one to the other is straightforward and
%  automatically performed when appropriate.  For example, \code{c(1, 2)[1.9]
%    == 1}.  This distinction is mainly just relevant to the implementation,
%  as integers and doubles are stored and dealt with differently in the
%  runtime but programmers only see the distinction when printing.
%
%\item \{\D, \l\} and \{\C, \l\}: at first glance, these type signatures are
%  suspect.  Why would so many programmers want to pass a double or a list to
%  some function?  The key insight here is that doubles and characters are
%  {\it vectorized} in R, meaning that a double is really a vector of
%  doubles.  Thus, these seemingly strange signatures are revealed to be
%  entirely reasonable: these signatures are likely to describe arguments
%  which are either vectors or lists.
%
%\item \{\ANY,\sN\}: in most languages, NULL is an inhabitant of all types,
%  and we see that this is not the case in R.  That said, there is a pattern
%  of many functions having default or optional arguments, and in many of
%  these cases NULL is the default value.
%
%\item \{\D,\L\}: the first values of type logical that come to mind are TRUE
%  and FALSE, which does little to explain this strange signature.  Indeed,
%  the key here is an implementation detail of R's: recall that NAs inhabit
%  every type.  As it happens, the default type of NA is logical (e.g.,
%  \code{typeof(NA) == "logical"}), and this is likely what is being seen
%  here.  We will control for this by distinguishing NAs with their own type,
%  {\tt raw\_NA}, but we will get to that in time.
%
%\item \{\C,\D\}, \{\C,\I\} and \{character, double, integer\}: yet more
%  strange signatures.  These signatures are a little more subtle to explain.
%  For one, indexing lists and vectors can be done with either strings (i.e.,
%  to access a named index) or with numbers.  Another option is dates: R has
%  many ways to present dates, and strings and numbers feature among them.
%  Finally, many \AT{get numbers} functions in R return a string if an error
%  occurs during execution.
%
%\end{enumerate}
%
%How will we deal with this?
%
%\begin{enumerate}
%\item consolidating doubles and integers into a single real-number type
%  would account for all of this polymorphism.
%\item to get an idea of how often vectors and lists co-occur in signatures,
%  we need to expand our analysis of vectorized types.  Notably, when we
%  encounter e.g. a double, we should collect length information to see if
%  the double is a vector of doubles or just a scalar.  This will feature in
%  our next type system, L1, which we will get to momentarily.
%\item if we roll NULLs into every type, we can cleanly deal with these X,
%  NULL signatures.  As we have established that an annotation for an
%  optional argument is useful, this will allow us to reveal other
%  interesting patterns.
%\item similarly with NULLs, we're interested in rolling NAs in to all types.
%  As we mentioned, this is actually what the R runtime does, but a default
%  type needed to be chosen for NAs appearing without context.  Subsuming NAs
%  should help to reveal other interesting patterns.
%\item dealing with this polymorphism will be more tricky.  We might need to
%  pull class information in (to pick up on dates), and possibly make a
%  signature judgment depending on which argument position has the signature
%  (e.g. a \{character, X\} signature on a return).  We will come back to
%  this later.
%
%\end{enumerate}

There are some clear limitations to the L0 type system that we should
address before working through any of these other issues. As we mentioned,
\code{typeof} makes no distinction between vectors and scalars, as indeed
there is no such distinction in R.  But as we saw, we might benefit from
such a distinction, and indeed an all-new runtime environment design for R
might benefit from e.g. not needing to vectorize scalar values.  This is only one
of the modifications we make, all of which are thoroughly explored in
the following section.
%
%
%
%
\subsection{L1}

As we mentioned in \AT{an earlier section}, and as was revealed in the previous section,
L0 is somewhat limited in its expressivity. To address this, we develop L1, which espouses
a finer-grained approach to types. Major modifications from L0 include the differentiation of
scalars and vectors, the parameterization of list types over element types, and the use of
attribute and class information to distinguish types for data frames and matrices.

@@

\begin{tabular}{@{}r|l|l|l@{}}\hline
\bf Argument & \bf Type & \bf Class &\bf Attributes \\
x & \sD & numeric & \{\} \\
y & \D & numeric & \{\} \\
retv & \lT{D} & list & \{\xspace example: \sC\}
\end{tabular}
\caption{Example call and L1 signature.}\label{fig:exL1}\end{figure}
